---
title: "Association Between Smoking and Mental Health: Empirical Evidence from IPUMS NHIS Data of US"
author: "Group Q: 
Haoxuan Li
Jing Liu
Lingjia Kang
Qiannuo Chen
Sitong Yan
Xuyuan Jin"
date: "2025.02.01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages

```{r}
# Load required packages
# This ensures that all necessary libraries for data manipulation, visualization, and statistical modeling are installed and loaded.
packages <- c("dplyr", "ggplot2", "tidyr", "corrplot", "ggcorrplot", "car", "survey", "ipumsr")
library(lmtest)
library(sandwich)
library(lmtest)
package.check <- lapply(packages, function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})
```

## Load and Clean Data

```{r}
# Load dataset using IPUMS
# The dataset contains various demographic and health-related variables, which will be used for statistical analysis.
ddi <- read_ipums_ddi("nhis_00005.xml")
data <- read_ipums_micro(ddi)
```

```{r}
# Select relevant variables and clean the data
# This step filters out inappropriate values and ensures that only adults (18+) are included.
data_clean <- data %>%
  dplyr::select(AGE, SEX, PSU, STRATA, PERWEIGHT, INCFAM07ON, INCFAM97ON2, HEALTH, EDUCREC1, NCHILD, EMPSTATIMP1, CIGDAYMO, CIGSDAY, HRSLEEP, SLEEPFALL, SLEEPSTAY, 
                AHOPELESS, ANERVOUS, ARESTLESS, ASAD, AWORTHLESS, AEFFORT) %>%
  filter(AGE >= 18, CIGSDAY <= 20, INCFAM07ON < 90, SLEEPSTAY < 90, SLEEPFALL < 90,
         AHOPELESS < 6, ANERVOUS < 6, ARESTLESS < 6, ASAD < 6, AWORTHLESS < 6, AEFFORT < 6, HEALTH < 6, EMPSTATIMP1 != 0, INCFAM07ON < 96, INCFAM97ON2 < 97)
```

## Feature Engineering: Creating a New Variable K6

```{r}
# Create a composite psychological distress score (K6) and adjust weight variable
# The K6 variable aggregates six mental health indicators to measure psychological distress.
data_clean <- data_clean %>%
  mutate(K6 = AHOPELESS + ANERVOUS + ARESTLESS + ASAD + AWORTHLESS + AEFFORT, 
         PERWEIGHT = PERWEIGHT / 4)  # Adjusting weight by dividing by 4
```

## Correlation Analysis

```{r}
# Calculate correlation between daily cigarette consumption and psychological distress
cor_test <- cor.test(data_clean$CIGSDAY, data_clean$K6, use = "complete.obs")
print(cor_test)
```

## Visualization: Relationship between K6 and Cigarette Consumption

```{r}
# Scatter plot with linear regression line
# This plot helps visualize the relationship between smoking and psychological distress.
ggplot(data_clean, aes(x = CIGSDAY, y = K6)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Cigarettes Per Day vs. Psychological Distress (K6)",
       x = "Cigarettes per Day", y = "K6 Score") +
  theme_minimal()
```

## Weighted Regression Analysis

```{r}
# Define survey design for weighted analysis
# This accounts for survey sampling weights to provide more accurate estimates.
design <- svydesign(ids = ~1, weights = ~PERWEIGHT, data = data_clean)
```

```{r}
# Perform weighted regression analysis
weighted_model <- svyglm(K6 ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON, design = design)
summary(weighted_model)
```

## F-test for Model Significance

```{r}
# Perform F-test to check the significance of predictors
f_test <- regTermTest(weighted_model, ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON)
print(f_test)
```

## Compute Weighted R-Squared and Adjusted R-Squared
```{r}
# Compute R-squared and adjusted R-squared for the weighted model
y_hat <- predict(weighted_model, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R2_weighted <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(weighted_model)) - 1  
adjusted_R2_weighted <- 1 - ((1 - R2_weighted) * (n - 1) / (n - p - 1))
print(R2_weighted)
print(adjusted_R2_weighted)
```

## Heteroscedasticity Tests

```{r}
# 1. Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(weighted_model)
print(bp_test)
```

```{r}
# 2. White test for heteroscedasticity (alternative method)
robust_se <- vcovHC(weighted_model, type = "HC0")  
robust_model <- coeftest(weighted_model, vcov = robust_se)
print(robust_model)
#the robust_mode was over-adjusted，therefore  the Robust Standar Errors: Design-Based Standard Errors was used as the final print result
summary(weighted_model, vartype = c("se", "ci"))
```

## Residual Plot for Heteroscedasticity Check
```{r}
# 3. Residual Plot
# Calculating the residual of svyglm
data_clean$residuals_svyglm <- residuals(weighted_model)
data_clean$fitted_svyglm <- fitted(weighted_model)

ggplot(data_clean, aes(x = fitted_svyglm, y = residuals_svyglm)) +
  geom_point(alpha = 0.5, color = "blue") + 
  geom_smooth(method = "loess", color = "red", se = FALSE) +  
  labs(title = "Residual Plot for Survey Weighted Generalized Linear Model",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

## Gender-Specific Analysis
```{r define-survey-design}
# Define survey design objects for males and females
# Purpose: To account for survey weights in the regression analysis.
data_male <- subset(data_clean, SEX == 1)
data_female <- subset(data_clean, SEX == 2)
design_male <- svydesign(ids = ~1, weights = ~PERWEIGHT, data = data_male)
design_female <- svydesign(ids = ~1, weights = ~PERWEIGHT, data = data_female)
```

```{r weighted-regression-male}
# Run weighted regression for males
# Purpose: To estimate regression coefficients while accounting for survey weights.
weighted_model_male <- svyglm(K6 ~ CIGSDAY + AGE + HEALTH + NCHILD + INCFAM07ON, 
                              design = design_male)
```

```{r weighted-regression-female}
# Run weighted regression for females
weighted_model_female <- svyglm(K6 ~ CIGSDAY + AGE + HEALTH + NCHILD + INCFAM07ON, 
                                design = design_female)
```

## Multicollinearity Check Using Variance Inflation Factor (VIF)
```{r vif-check-male}
# Check for multicollinearity in the male model
# Purpose: To detect whether predictor variables are highly correlated.
vif(weighted_model_male)
```

```{r vif-check-female}
# Check for multicollinearity in the female model
vif(weighted_model_female)
```

## Model Fit Assessment

```{r f-test-male}
# Compute F-test for the overall significance of the male model
# Purpose: To test if at least one predictor variable significantly contributes to the model.
f_test_male <- regTermTest(weighted_model_male, ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON)
print(f_test_male)
```

```{r f-test-female}
# Compute F-test for the female model
f_test_female <- regTermTest(weighted_model_female, ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON)
print(f_test_female)
```

```{r r-squared-male}
# Compute R-squared and Adjusted R-squared for the male model
# Purpose: To measure how well the regression model explains the variance in the dependent variable.
y_hat_male <- predict(weighted_model_male, type = "response")
y_male <- design_male$variables$K6  
w_male <- weights(design_male)  
sst_male <- sum(w_male * (y_male - weighted.mean(y_male, w_male))^2)
sse_male <- sum(w_male * (y_male - y_hat_male)^2)
R2_male <- 1 - (sse_male / sst_male)
n_male <- nrow(design_male$variables) 
p_male <- length(coef(weighted_model_male)) - 1  
adjusted_R2_male <- 1 - ((1 - R2_male) * (n_male - 1) / (n_male - p_male - 1))
print(R2_male)  
print(adjusted_R2_male)  
```

```{r r-squared-female}
# Compute R-squared and Adjusted R-squared for the female model
y_hat_female <- predict(weighted_model_female, type = "response")
y_female <- design_female$variables$K6  
w_female <- weights(design_female)  
sst_female <- sum(w_female * (y_female - weighted.mean(y_female, w_female))^2)
sse_female <- sum(w_female * (y_female - y_hat_female)^2)
R2_female <- 1 - (sse_female / sst_female)
n_female <- nrow(design_female$variables) 
p_female <- length(coef(weighted_model_female)) - 1  
adjusted_R2_female <- 1 - ((1 - R2_female) * (n_female - 1) / (n_female - p_female - 1))
print(R2_female) 
print(adjusted_R2_female) 
```

## Heteroscedasticity Test
```{r bp-test-male}
# Perform Breusch-Pagan test for heteroscedasticity in the male model
# Purpose: To check if variance of residuals is constant across all fitted values.
bp_test_male <- bptest(weighted_model_male) 
print(bp_test_male)
```
```{r bp-test-female}
# Perform Breusch-Pagan test for the female model
bp_test_female <- bptest(weighted_model_female) 
print(bp_test_female)
```

## Weighted Regression Analysis and Robustness Checks

### 1. Standard Errors with PSU Clustering

This section calculates the standard errors (SE) and confidence intervals (CI) for the weighted regression models separately for male and female groups, considering PSU clustering.

```{r standard_errors}
# Compute standard errors and confidence intervals for male model
summary(weighted_model_male, vartype = c("se", "ci"))

# Compute standard errors and confidence intervals for female model
summary(weighted_model_female, vartype = c("se", "ci"))
```

### 2. Residual Analysis

Residual analysis helps assess model assumptions such as homoscedasticity and normality. Here, we extract fitted values and residuals for both male and female models and visualize them using residual plots.

```{r residual_analysis}
# Extract fitted values and residuals for male model
residuals_male <- data.frame(
  Fitted_Values = fitted(weighted_model_male),
  Residuals = residuals(weighted_model_male)
)

# Extract fitted values and residuals for female model
residuals_female <- data.frame(
  Fitted_Values = fitted(weighted_model_female),
  Residuals = residuals(weighted_model_female)
)

# Plot residuals for male model
ggplot(residuals_male, aes(x = Fitted_Values, y = Residuals)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  ggtitle("Residual Plot for Male (Weighted Regression)") +
  xlab("Fitted Values") + ylab("Residuals")

# Plot residuals for female model
ggplot(residuals_female, aes(x = Fitted_Values, y = Residuals)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  ggtitle("Residual Plot for Female (Weighted Regression)") +
  xlab("Fitted Values") + ylab("Residuals")
```

#===========================================================================
#===========================================================================

###Robustness Checks

To ensure the robustness of our regression results, we run multiple models with increasing levels of control variables.

#### Model 1: Core Variable Only

This model includes only the core independent variable (CIGSDAY).

```{r model_1}
# 1️: Only considering the core independent variable
model_1 <- svyglm(K6 ~ CIGSDAY, design = design)
summary(model_1, vartype = c("se", "ci"))
# Calculating the F-Statistic
f_test <- regTermTest(model_1, ~ CIGSDAY)
print(f_test) 
# R-squared & adjusted
y_hat <- predict(model_1, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R21 <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(model_1)) - 1  
adjusted_R21 <- 1 - ((1 - R21) * (n - 1) / (n - p - 1))
print(R21)
print(adjusted_R21)
```

#### Model 2: Adding Demographic Variables

This model includes age and sex as additional covariates.

```{r model_2}
# 2️:Adding the variables of age and sex
model_2 <- svyglm(K6 ~ CIGSDAY + AGE + SEX, design = design)
summary(model_2, vartype = c("se", "ci"))
# Calculating the F-Statistic
f_test <- regTermTest(model_2, ~ CIGSDAY+ AGE + SEX)
print(f_test) 
# R-squared & adjusted
y_hat <- predict(model_2, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R22 <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(model_2)) - 1  
adjusted_R22 <- 1 - ((1 - R22) * (n - 1) / (n - p - 1))
print(R22)
print(adjusted_R22)
```

#### Model 3: Adding Health and Family Variables

This model adds health status and number of children as additional covariates.

```{r model_3}
# 3️: Adding variables about general health condition and family
model_3 <- svyglm(K6 ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD, design = design)
summary(model_3, vartype = c("se", "ci"))
# Calculating the F-Statistic
f_test <- regTermTest(model_3, ~ CIGSDAY+ AGE + SEX + HEALTH + NCHILD)
print(f_test) 
# R-squared & adjusted
y_hat <- predict(model_3, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R23 <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(model_3)) - 1  
adjusted_R23 <- 1 - ((1 - R23) * (n - 1) / (n - p - 1))
print(R23)
print(adjusted_R23)
```

#### Model 4: Adding Economic Variables

This model further incorporates family income as an additional control variable.

```{r model_4}
# 4️: Adding income related variable
model_4 <- svyglm(K6 ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON, design = design)
summary(model_4, vartype = c("se", "ci"))
# Calculating the F-Statistic
f_test <- regTermTest(model_4, ~ CIGSDAY+ AGE + SEX + HEALTH + NCHILD + INCFAM07ON)
print(f_test) 
# R-squared & adjusted
y_hat <- predict(model_4, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R24 <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(model_4)) - 1  
adjusted_R24 <- 1 - ((1 - R24) * (n - 1) / (n - p - 1))
print(R24)
print(adjusted_R24)

```

#### Model 5: Full Model Including Sleep Variables

This is the most comprehensive model, incorporating sleep-related variables as well.

```{r model_5}
# 5️: Adding sleep related variable (entire model)
model_5 <- svyglm(K6 ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON + SLEEPFALL + SLEEPSTAY, design = design)
summary(model_5, vartype = c("se", "ci"))
# Calculating the F-Statistics for the entire model
f_test <- regTermTest(model_5, ~ CIGSDAY + AGE + SEX + HEALTH + NCHILD + INCFAM07ON + SLEEPFALL + SLEEPSTAY)
print(f_test) 
# R-squared & adjusted
y_hat <- predict(model_5, type = "response")
y <- data_clean$K6
w <- weights(design)
sst <- sum(w * (y - weighted.mean(y, w))^2)
sse <- sum(w * (y - y_hat)^2)
R25 <- 1 - (sse / sst)
n <- nrow(data_clean) 
p <- length(coef(model_5)) - 1  
adjusted_R25 <- 1 - ((1 - R25) * (n - 1) / (n - p - 1))
print(R25)
print(adjusted_R25)
```
#View the mediating effects of sleep-related variables
```{r mediate_model_4}
model_sc <- svyglm(SLEEPFALL ~ CIGSDAY, design = design, family = gaussian())
model_ks <- svyglm(K6 ~ CIGSDAY + SLEEPFALL, design = design, family = gaussian())
mediate_model_3 <- mediate(model_sc, model_ks, treat = "CIGSDAY", mediator = "SLEEPFALL", boot = TRUE, sims = 1000)
summary(mediate_model_3)

model_ssc <- svyglm(SLEEPSTAY ~ CIGSDAY, design = design, family = gaussian())
model_kss <- svyglm(K6 ~ CIGSDAY + SLEEPSTAY, design = design, family = gaussian())
mediate_model_4 <- mediate(model_ssc, model_kss, treat = "CIGSDAY", mediator = "SLEEPSTAY", boot = TRUE, sims = 1000)
summary(mediate_model_4)

```

#Check whether the regression control variables affect CIGSDAY. Analyze all control variables one by one. If they are significant for CIGSDAY here and also significant for K6 in the basic regression model, they may be considered as confounding variables.
```{r model_conc}
model_conc <- svyglm(CIGSDAY ~ SLEEPFALL , design = design, family = gaussian())
summary(model_conc)
```


#age group 
```{r age_model}
univariate_model_group1<-data_clean%>%filter(AGE<=30)
univariate_model_group2<-data_clean%>%filter(AGE>30&AGE<=50)
univariate_model_group3<-data_clean%>%filter(AGE>50)
#Weighted regression for age group 1 (18-30)
design <- svydesign(ids = ~1, weights = ~PERWEIGHT, data = univariate_model_group1)
age_model <- svyglm(K6 ~ CIGSDAY + SEX + HEALTH + NCHILD + INCFAM07ON, 
                         design = design)
summary(age_model,vartype = c("se", "ci"))

```

### Conclusion

This analysis systematically examines the relationship between smoking (CIGSDAY) and psychological distress (K6) while incorporating various control variables in a stepwise manner. The robustness checks ensure the reliability of the results by evaluating the model fit and significance across different specifications.

